Explaining probabilistic prediction on the simplex with Shapley compositions

### 2023-12-14 paper/main.tex
edit paper, names in summa

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}
  \centering
  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
    }
  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
  }
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

@@

\subsection{Summarize explaination with angles, norms and projections}

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, this require a careful choose of the sequential binary partition and of the ILR composition to visualize. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction. The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory. The projection of a Shapley composition on the set of class composition informs in favor of, or against which, classes a feature is contribution. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation providing the norm of Shapley compositions, angles between them and their projection on class compositions.

@@

We know small expé..., sounds tricky,... first step for a theoretically founded multiclass problems explanations...
...

@@

\newpage
\section{Class compositions}

A $k$-class compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
  \end{aligned}
\end{equation}
To summarize, the $i$th part of a $k$-class compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
\begin{equation}
    c_i^{(k)} =
  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
     1, & \text{if}\ i=k \\
     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
  \end{array}\right.\right).
\end{equation}
In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.

\newpage
\section{Summarizing the explanation with norms, angles and projections of Shapley compositions}

TODO...

>>>>

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}
  \centering
  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
    }
  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
  }
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

@@

\subsection{Angles, norms and projections}

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction. The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.

@@

We know small expé..., sounds tricky,... first step for a theoretically founded multiclass problems explanations...
...
Features INDEPENDENCE!!

Axiomatic formulation

@@

\newpage
\section{Class-compositions}

A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
  \end{aligned}
\end{equation}
To summarize, the $i$th part of a $k$-class-compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
\begin{equation}
    c_i^{(k)} =
  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
     1, & \text{if}\ i=k \\
     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
  \end{array}\right.\right).
\end{equation}
In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.

\newpage
\section{Summarizing the explanation with norms, angles and projections of Shapley compositions}
\label{app:summarize}

The example of Figure \ref{fig:3classes} can be summarized by computing the norms of the Shapley compositions, angles between them and their projection on the set of class-compositions.
